---
title: "Star Wars Lab"
output: html_notebook
---

# Learning Objectives:

- Discuss Things
- Learn about notebooks
- What is a data frame?
    - Rows V columns
    - What is a row of data?
    - Data types
- Tour of the RStudio Cloud interface
- How to run a code chunk
- How to render a notebook
- Structure of a R command (function)


# Discuss Things

- What was (is) your reaction to knowing I would use the survey results in class?
- How do you feel about Governor Hochul's changes to defined races in New York?
- 



# Notebooks

Our work will be managed in documents called notebooks. This document
you are viewing is a notebook. A RMarkdown notebook combines plain
text with R code to create a reproducible document.

The lecture notes are a notebook. The lab is a notebook. The work we
do for this class will always be in a notebook. I will make R scripts
available for transparency when I hide things from you.

In the data folder, there is a R script I wrote to scrub PII from the
class survey.


# Setup

- To make sure you have everything loaded you will need for the
  lecture/lab, look for a section labeled "Setup" and run the code
  block in that section of the notebook.
- Run the code chunk now, PLEASE!!!
    - Use the little green arrow to the right of the editor window.
    - Or use the keyboard shortcut, `Ctrl+Shift+Enter`

```{r}
library(tidyverse)
library(ggrepel)
data(starwars)
```

- A R function is often a verb, such as `import()`.
- Some functions, like `library()` are not verbs. 🤷
- Immediately after a function is:
    - an "open parenthesis",
    - followed by one or more "parameters",
    - followed by a "close parenthesis".
- R code is composed of lots of "little" functions which are often chained together.
    - This is also called functional composition. For our purposes, these two terms have basically the same meaning.
    - Very different from SAS/SPSS syntax which is based around a smaller set of functions which are more powerful.


# Data

```{r}
starwars
```

- Observe how the results of the code are embedded in the notebook.
    - When you run the code in a notebook, R creates a new file,
      `lab.nb.html`, which contains the original code AND the HTML
      containing the output.
    - In a healthcare setting, this `nb.html` file must be carefully
      protected because it can contain PHI.
    - A .Rmd file only contains PHI if you enter it manually into the
      text or code, and why on earth would you do that?
- `starwars` is a data frame
- What does a single row of data represent?
- What kinds of data types do we have?

le

# Plotting

- Our first few lectures/labs are focused on plotting/graphing data.
- To run a single line of code within a code chunk, `Ctrl+Enter`.

```{r}
## Look!!!
## We are using TWO (2) functions to draw a plot!
## This is function chaining.
ggplot(data = starwars, mapping = aes(x = homeworld)) +
    geom_bar()
```

- **Oops:** I can't read the X Axis can you?
- Notice how RStudio superimposes formatting on our markdown.



```{r}
ggplot(
    data = starwars, mapping = aes(y = homeworld)) +
    geom_bar()
```

Much better!

```{r}

# Can YOU draw a bar plot showing the number of characters by species?

```

```{r}
ggplot(
    data = starwars, mapping = aes(x = mass, y = height)) +
    geom_point()
```

We've got ONE REALLY heavy character.


```{r}
ggplot(
    data = starwars, mapping = aes(x = mass, y = height, label = name)) +
    geom_point() +
    geom_text_repel()
```

Damn Hutts. Drop all beings which weigh more than 1,000 pounds from
our analysis.

```{r}
lighter_starwars <-
    starwars %>%
    filter(mass < 1000)

ggplot(
    data = lighter_starwars,
    mapping = aes(x = mass, y = height, label = name)
) +
    geom_point() +
    geom_text_repel()
```

And now a basic regression line.

```{r}
ggplot(
    data = lighter_starwars,
    mapping = aes(x = mass, y = height, label = name)
    ) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE) +
    geom_text_repel()
```

Last challenge for tonight:

```{r}
## - Load the midwest data set.
## - What is the relationship between college education and poverty?
## - Hint: Look at the percollege and percbelowpoverty
data(midwest)
?midwest

## YOUR TURN!

```
