---
title: "Bechdel"
author: "Your Name Here"
output: html_notebook
---

In this mini analysis we work with the data used in the
FiveThirtyEight story titled ["The Dollar-And-Cents Case Against
Hollywood's Exclusion of
Women"](https://fivethirtyeight.com/features/the-dollar-and-cents-case-against-hollywoods-exclusion-of-women/).
Your task is to fill in the blanks denoted by `___`.



# Setup

- Observe how this lab combines R syntax with regular language.
- The R syntax parts are called code chunks. You can run one of these
  code chunks by pressing the green arrow below.
- To begin our analysis, we will first load some packages, to make R easier to use.
    - It is a best practice to always load any/all packages you need BEFORE the rest of your script/program.
    - You would then import/access your data, but in this case, the
      data is part of the `fivethirtyeight` pacakage.
- For more information on the Bechdel test, see [Wikipedia - Bechdel
  Test](https://en.wikipedia.org/wiki/Bechdel_test)



# Setup

```{r message=FALSE}
rm(list = ls())
## Loading packages, and why we do this first is something we will discuss in more detail later in the class.
library(fivethirtyeight)
library(tidyverse)
```

The data set contains information on `r nrow(bechdel)` movies released
between `r min(bechdel$year)` and `r max(bechdel$year)`.  However
we'll focus our analysis on movies released between 1990 and 2013.

```{r}
bechdel90_13 <- 
  bechdel %>% 
  filter(between(year, 1990, 2013))
```

We use the R terminal to perform analyses not needed in our final
output.

Below: we want to replace the `___` with some code. Look in the
previous code/paragraphs for inspiration. R's syntax may be foreign to
you, but functions do things and they are usually named clearly, to
help you make use of them.

There are `___` such movies.

The financial variables we'll focus on are the following:

-   `budget_2013`: Budget in 2013 inflation adjusted dollars
-   `domgross_2013`: Domestic gross (US) in 2013 inflation adjusted dollars
-   `intgross_2013`: Total International (i.e., worldwide) gross in 2013 inflation adjusted dollars

And we'll also use the `binary` and `clean_test` variables for
**grouping**.

## Analysis

Let's take a look at how median budget and gross vary by whether the
movie passed the Bechdel test, which is stored in the `binary`
variable.

```{r message = FALSE}
## Notice how this changes the GRAIN of the data.
bechdel90_13 %>%
  group_by(binary) %>%
  summarise(med_budget = median(budget_2013),
            med_domgross = median(domgross_2013, na.rm = TRUE),
            med_intgross = median(intgross_2013, na.rm = TRUE))
```

Question: Why are we using median rather than average?

Next, let's take a look at how median budget and gross vary by a more
detailed indicator of the Bechdel test result.  This information is
stored in the `clean_test` variable, which takes on the following
values:

-   `ok` = passes test
-   `dubious`
-   `men` = women only talk about men
-   `notalk` = women don't talk to each other
-   `nowomen` = fewer than two women

```{r message = FALSE}
bechdel90_13 %>%
  #group_by(___) %>%
  summarise(med_budget = median(budget_2013),
            med_domgross = median(domgross_2013, na.rm = TRUE),
            med_intgross = median(intgross_2013, na.rm = TRUE))
```

In order to evaluate how return on investment varies among movies that
pass and fail the Bechdel test, we'll first create a new variable
called `roi` as the ratio of the gross to budget.

```{r}
bechdel90_13 <- bechdel90_13 %>%
  mutate(roi = (intgross_2013 + domgross_2013) / budget_2013)
```

Let's see which movies have the highest return on investment.

```{r}
bechdel90_13 %>%
  arrange(desc(roi)) %>% 
  select(title, roi, year)
```

Below is a visualization of the return on investment by test result,
however it's difficult to see the distributions due to a few extreme
observations.

```{r warning = FALSE}
ggplot(data = bechdel90_13, 
       mapping = aes(x = clean_test, y = roi, color = binary)) +
  geom_boxplot() +
  labs(title = "Return on investment vs. Bechdel test result",
       x = "Detailed Bechdel result",
       y = "___",
       color = "Binary Bechdel result")
```

What are those movies with *very* high returns on investment?

```{r}
bechdel90_13 %>%
  filter(roi > 400) %>%
  select(title, budget_2013, domgross_2013, year)
```

Zooming in on the movies with `roi < ___` provides a better view of
how the medians across the categories compare:

```{r warning = FALSE}
ggplot(data = bechdel90_13, mapping = aes(x = clean_test, y = roi, color = binary)) +
  geom_boxplot() +
  labs(title = "Return on investment vs. Bechdel test result",
       subtitle = "___", # Something about zooming in to a certain level
       x = "Detailed Bechdel result",
       y = "Return on investment",
       color = "Binary Bechdel result") +
  coord_cartesian(ylim = c(0, 15))
```

Thank you for participating in your first lab. You will all get full
credit for this lab simply by going into Canvas and marking the lab
complete.
