---
title: "Star Wars Lab"
format: html
editor: 
  markdown: 
    wrap: 72
---

# Learning Objectives:

1. Have some fun.
2. Get into our lab groups.
3. Tour of the RStudio Cloud interface
4. Learn about notebooks
  - Code vs Visual Render
5. What is a data frame?
  - Rows V columns
  - What is a row of data?
  - Data types
6. How to run a code chunk
7. How to render an entire notebook
8. Global Environment and how to clear it (sweep).
  - Switching from one document to another.
9. Structure of a R command (function)
10. (15 minutes) Each group needs to come up with one topic they would like explore during this class, at least briefly. This is not a promise, but it is interesting to explore data that interests you. Look for a topic which we can use to explore data and answer questions and for which there is a data set.


And, honestly, that's enough for one night.

# Notebooks

Our work will be managed in documents called notebooks. This document you are viewing is a notebook. A RMarkdown notebook combines plain text with R code to create a reproducible document.

The lecture notes are a notebook. The lab is a notebook. The work we do for this class will always be in a notebook. I will make all R scripts used in this course available for you to study/review.

# Setup

- To make sure you have everything loaded you will need for the lecture/lab, look for a section labeled "Setup" and run the code block in that section of the notebook.
- Run the code chunk now, PLEASE!!!
  - Use the little green arrow to the right of the editor window.
  - Or use the keyboard shortcut, `Ctrl+Shift+Enter`

```{r}
#| echo: false

# Loads two packages into our current session.
library(tidyverse)
library(ggrepel)

# Loads Star Wars data into our current session.
data(starwars)
```

- A R function is often a verb, such as `import()`.
- Some functions, like `library()` are not verbs. 🤷
- Immediately after a function is:
  - an "open parenthesis",
  - followed by one or more "parameters",
  - followed by a "close parenthesis".
- R code is composed of lots of "little" functions which are often chained together.
  - This is also called functional composition. For our purposes, these two terms have basically the same meaning.
  - Very different from SAS/SPSS syntax which is based around a smaller set of functions which are more powerful.

# Data

```{r}
# This prints the data into our notebook.
starwars
```

- Observe how the results of the code are embedded in the notebook.
  - When you run the code in a notebook, R creates a new file, `lab.nb.html`, which contains the original code AND the HTML containing the output.
  - In a healthcare setting, this `nb.html` file must be carefully protected because it can contain PHI.
  - A .qmd file only contains PHI if you enter it manually into the text or code, and why on earth would you do that?
- `starwars` is a data frame
- What does a single row of data represent?
- What kinds of data types do we have?

# Plotting

- Our first few lectures/labs are focused on plotting/graphing data.
- A plot is a graphical form of summarization.
  - From many data points to a single plot.
  - Helps us focus on the patterns, not the individual details.
- To run a single line of code within a code chunk, `Ctrl+Enter`.

```{r}
## Look!!!
## We are using TWO (2) functions to draw a plot!
## This is a form of function chaining/composition.
ggplot(data = starwars, mapping = aes(x = homeworld)) +
    geom_bar()
```

- **Oops:** I can't read the X Axis can you?
- Notice how RStudio superimposes formatting on our markdown.

```{r}
ggplot(
    data = starwars, mapping = aes(y = homeworld)) +
    geom_bar()
```

Much better!

```{r}
# Can YOU draw a bar plot showing the number of characters by species?

```

```{r}
ggplot(
    data = starwars, mapping = aes(x = mass, y = height)) +
    geom_point()
```

We've got ONE REALLY heavy character.

```{r}
ggplot(
    data = starwars, mapping = aes(x = mass, y = height, label = name)) +
    geom_point() +
    geom_text_repel()
```

We can (always) blame the Hutts. Drop all beings which weigh more than 1,000 pounds from our analysis.

```{r}
lighter_starwars <-
    starwars |>
    filter(mass < 1000)

ggplot(
    data = lighter_starwars,
    mapping = aes(x = mass, y = height, label = name)
) +
    geom_point() +
    geom_text_repel()
```

Add a basic regression line.

```{r}
ggplot(
    data = lighter_starwars,
    mapping = aes(x = mass, y = height, label = name)
    ) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE) +
    geom_text_repel()
```

Last challenge for tonight:

```{r}
## - Load the midwest data set.
## - What is the relationship between college education and poverty?
## - Hint: Look at the percollege and percbelowpoverty
data(midwest)
?midwest

## YOUR TURN!

```
